---
title: "AML_Lab3_Hyungyum_Kim"
author: "Josh Hyungyum Kim"
date: "2018 - 10 - 4"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Implement the SSM. 

```{r}

# 1.

# T = 100
Tsteps <- 100

# Vectors for true states and observations
Tstates <- c()
obs <- c()

# Initial model
set.seed(123456) # seed for reproducible results
init <- round(runif(1, 0, 100))

# simulation of 100 time steps
set.seed(123456)
for(i in 1:Tsteps){
  
  if (i==1){
    Tstates[1] <- init
  }else{
    # select one model of mixtures
    mix_comp2 <- sample((0:2),1)
    Tstates[i] <- round(rnorm(1, Tstates[i-1]+mix_comp2, 1))  
  }
  
  # select one model of mixtures
  mix_comp <- sample((-1:1),1)
  # observation 
  obs[i] <- round(rnorm(1, Tstates[i]+mix_comp, 1))
}

# Print the results
list("True_location"=Tstates, "Expected_location"=obs)
```

The SSM described in lab instruction is implemented and the simulation of 100 time steps are listed above. The observations will be used to identify the state via particle filtering(100 particles). Particle filtering algorithm is implemented as below:



```{r, fig.height=6}

# particle filter

# transition model
trans_model <- function(z){
  iter <- length(z)
  res <- c()
  
  for(i in 1:iter){
    mix_comp <- sample((0:2), 1)
    res[i] <- rnorm(1, z[i]+mix_comp, 1)
  }
  res
}

# density function
den_fun <- function(mean, input, sd){
  (dnorm(x=input, mean=mean, sd=sd)+
   dnorm(x=input, mean=mean+1, sd=sd)+
   dnorm(x=input, mean=mean-1, sd=sd))/3
}

# 100 particles
num_parti <- 100

particle <- function(weight=TRUE, sd=1){
  
  # lists for storing weights and particles
  Zn <- list()
  Wn <- list()
  
  # 100 samples
  Zn[[1]] <- runif(num_parti, 0 ,100)
  
  for(i in 1:Tsteps){
    
    numer <- sapply(X = Zn[[i]], FUN = den_fun, input=obs[i], sd=sd)
    Wn[[i]] <- numer / sum(numer)
    
    if(i==Tsteps){break}
    
    # correction and prediction
    if(weight==TRUE){
      ind <- sample(1:num_parti, 100, replace=TRUE, prob=Wn[[i]])
    }else{
      ind <- sample(1:num_parti, 100, replace=TRUE)
    }
    
    # new particle
    Zn[[i+1]] <- trans_model(Zn[[i]][ind])
  }
  list("Weights"=Wn, "Particles"=Zn)
}

res <- particle()

parti_plot <- function(res){
  Wn <- res$Weights
  Zn <- res$Particles
  
  par(mfrow=c(2,2))
  for(i in c(1,26,50,100)){
    hist(Zn[[i]], freq = FALSE, col="grey88", breaks=10,
         main=i, xlab="Position")
    lines(density(Zn[[i]]), col="lightblue")
    abline(v=sum(Wn[[i]]*Zn[[i]]), col="blue", lwd=2) # Prediction
    abline(v=Tstates[i], col="Red", lwd=2) # True State
  }
}

parti_plot(res)
```

Above 4 different histograms show particles of time step 1, 26, 50 and 100. Lightblue line means approximated density, blue line means expected location and the red line means the true location. 

\clearpage

### 2. Repeat with SD=5, 50.

```{r, fig.height=6}

# 2.

# simulation of 100 time steps with sd=5
set.seed(123456)
for(i in 1:Tsteps){
  
  # select one model of mixtures
  mix_comp <- sample((-1:1),1)
  # observation 
  obs[i] <- round(rnorm(1, Tstates[i]+mix_comp, 5))
}

# result with SD=5
res5 <- particle(sd=5)
parti_plot(res5)
title("Result with SD=5", outer=TRUE, line=-1)


# simulation of 100 time steps with sd=50
set.seed(123456)
for(i in 1:Tsteps){
  
  # select one model of mixtures
  mix_comp <- sample((-1:1),1)
  # observation 
  obs[i] <- round(rnorm(1, Tstates[i]+mix_comp, 50))
}

# result with SD=50
res50 <- particle(sd=50)
parti_plot(res50)
title("Result with SD=50", outer=TRUE, line=-1)
```
Based on the above plots, one might say that it is obvious the particles become more and more sparsed when the sd increases. This is a natural outcome since the sd influence the probability distributions of emission model. Furthermore, the results of prediction get worse since high standard deviation gives more uncertatinty for observations. 

\clearpage

### 3. Weight=1.


```{r, fig.height=6}

# 3. 

# To use same True state and obs for Q3 as Q1
set.seed(123456)
for(i in 1:Tsteps){
  
  if (i==1){
    Tstates[1] <- init
  }else{
    # select one model of mixtures
    mix_comp2 <- sample((0:2),1)
    Tstates[i] <- round(rnorm(1, Tstates[i-1]+mix_comp2, 1))  
  }
  
  # select one model of mixtures
  mix_comp <- sample((-1:1),1)
  # observation 
  obs[i] <- round(rnorm(1, Tstates[i]+mix_comp, 1))
}

res_nofil <- particle(weight = FALSE)
parti_plot(res_nofil)
title("Result with weight=1", outer=TRUE, line=-1)

```

When the weight=1 for every particle which means there is no correction, results of 4 time steps are really close to the true states. Maybe it is because the transition model is the same and the particles could move along with the observations. However, the histograms of particles are very sparsed  than any weighted runs above.






```{r, echo=FALSE, eval=FALSE}

# 1.

# seed for reproducible results
set.seed(12345)

# T = 100
Tsteps <- 100

# Initial model
init <- round(runif(1, 0, 100))

# Emission model
emi_model <- function(Zt){
  emi_res <- list()
  for(i in 1:length(Zt)){
    mix_comp <- sample((-1:1),1)
    emi_res[[i]] <- rnorm(1, Zt[[i]]+mix_comp, 1)
  }
  emi_res
}

# Transition model
trans_model <- function(Zt_1){
  trans_res <- list()
  for(i in 1:length(Zt_1)){
    mix_comp <- sample((0:2),1)
    trans_res[[i]] <- rnorm(1, Zt_1[[i]]+mix_comp, 1)
  }
  trans_res
}

# Vectors for true states and observations
Tstates <- list()
obs <- list()

# Simulate it for 100 time steps
for(i in 1:Tsteps){
  if(i==1){
    obs[i] <- emi_model(init)
    Tstates[i] <- init
  }else{
    Tstates[i] <- trans_model(Tstates[[i-1]])
    obs[i] <- emi_model(Tstates[[i]])
  }
}

# Print the results
list("True_location"=round(unlist(Tstates)), "Expected_location"=round(unlist(obs)))

# Density calculation function
den_fun <- function(Xt, mean, sd){
  res <- ((dnorm(x=Xt, mean=mean, sd=sd) + dnorm(x=Xt, mean=mean-1, sd=sd) + dnorm(x=Xt, mean=mean+1, sd=sd))/3)
  res
}

# Number of particles
num_parti <- 100

# Particle function
particle <- function(sd=1){
  # Lists for storing  Zn and weights
  Zn <- list()
  Wn <- list()
  
  # Initialization
  Zn[[1]] <- runif(num_parti, 0, 100)
  
  for(i in 1:Tsteps){
    # Weights calculation // slide 10 of SSM_L3
    numer <- as.vector(sapply(X=Zn[[i]], FUN=den_fun, Xt=obs[[i]], sd=sd))
    Wn[[i]] <- numer / sum(numer)
    
    # Loop exit
    if(i==Tsteps){break}
    
    # Correction and prediction
    ind <- sample(1:num_parti, 100, replace=TRUE, prob=Wn[[i]])
    Zn[[i+1]] <- unlist(trans_model(Zn[[i]][ind]))
  }
  list("Weights"=Wn, "Particles"=Zn)
}

# Get results with sd=1
res <- particle(sd=1)

# Plot results
for(i in c(1,26,50,100)){
  
  plot(res$Particles[[i]], rep(1, num_parti), xlim=c(0,200))
  abline(v=sum(res$Weights[[i]]*res$Particles[[i]]), col="red")
}


# unlist(Tstates[c(1,26,50,100)])
# unlist(obs[c(1,26,50,100)])
# 
# sum(res$Weights[[1]]*res$Particles[[1]])
# sum(res$Weights[[26]]*res$Particles[[26]])
# sum(res$Weights[[50]]*res$Particles[[50]])
# sum(res$Weights[[100]]*res$Particles[[100]])
# 
# par(mfrow=c(2,2))
# 
# hist(res$Particles[[1]], freq=FALSE)
# abline(v=c(Tstates[[1]], sum(res$Weights[[1]]*res$Particles[[1]])), col=c(1,2), lwd=2)
# 
# hist(res$Particles[[26]], freq=FALSE)
# abline(v=c(Tstates[[26]], sum(res$Weights[[26]]*res$Particles[[26]])), col=c(1,2), lwd=2)
# 
# hist(res$Particles[[50]], freq=FALSE)
# abline(v=c(Tstates[[50]], sum(res$Weights[[50]]*res$Particles[[50]])), col=c(1,2), lwd=2)
# 
# 
# hist(res$Particles[[100]], freq=FALSE)
# abline(v=c(Tstates[[100]], sum(res$Weights[[100]]*res$Particles[[100]])), col=c(1,2), lwd=2)

```


```{r, echo=FALSE, eval=FALSE}

# 2.

# Emission model with sd=5
emi_model5 <- function(Zt, sd=5){
  emi_res <- list()
  for(i in 1:length(Zt)){
    emi_res[[i]] <- round((rnorm(1, Zt[[i]], 5) + rnorm(1, Zt[[i]]-1, 5) + rnorm(1, Zt[[i]]+1, 5))/3)
  }
  emi_res
}

# Simulate it for 100 time steps
for(i in 1:Tsteps){
  if(i==1){
    obs[i] <- emi_model5(init)
    Tstates[i] <- init
  }else{
    Tstates[i] <- trans_model(Tstates[[i-1]])
    obs[i] <- emi_model5(Tstates[[i]])
  }
}

res5 <- particle(sd=5)

unlist(Tstates[c(1,26,50,100)])
unlist(obs[c(1,26,50,100)])

sum(res5$Weights[[1]]*res5$Particles[[1]])
sum(res5$Weights[[26]]*res5$Particles[[26]])
sum(res5$Weights[[50]]*res5$Particles[[50]])
sum(res5$Weights[[100]]*res5$Particles[[100]])

par(mfrow=c(2,2))

hist(res5$Particles[[1]], freq=FALSE)
abline(v=c(Tstates[[1]], sum(res5$Weights[[1]]*res5$Particles[[1]])), col=c(1,2), lwd=2)

hist(res5$Particles[[26]], freq=FALSE)
abline(v=c(Tstates[[26]], sum(res5$Weights[[26]]*res5$Particles[[26]])), col=c(1,2), lwd=2)

hist(res5$Particles[[50]], freq=FALSE)
abline(v=c(Tstates[[50]], sum(res5$Weights[[50]]*res5$Particles[[50]])), col=c(1,2), lwd=2)


hist(res5$Particles[[100]], freq=FALSE)
abline(v=c(Tstates[[100]], sum(res5$Weights[[100]]*res5$Particles[[100]])), col=c(1,2), lwd=2)


# Emission model with sd=50
emi_model50 <- function(Zt, sd=50){
  emi_res <- list()
  for(i in 1:length(Zt)){
    emi_res[[i]] <- round((rnorm(1, Zt[[i]], 50) + rnorm(1, Zt[[i]]-1, 50) + rnorm(1, Zt[[i]]+1, 50))/3)
  }
  emi_res
}

# Simulate it for 100 time steps
for(i in 1:Tsteps){
  if(i==1){
    obs[i] <- emi_model50(init)
    Tstates[i] <- init
  }else{
    Tstates[i] <- trans_model(Tstates[[i-1]])
    obs[i] <- emi_model50(Tstates[[i]])
  }
}

res50 <- particle(sd=50)

unlist(Tstates[c(1,26,50,100)])
unlist(obs[c(1,26,50,100)])

sum(res50$Weights[[1]]*res50$Particles[[1]])
sum(res50$Weights[[26]]*res50$Particles[[26]])
sum(res50$Weights[[50]]*res50$Particles[[50]])
sum(res50$Weights[[100]]*res50$Particles[[100]])

par(mfrow=c(2,2))

hist(res50$Particles[[1]], freq=FALSE)
abline(v=c(Tstates[[1]], sum(res50$Weights[[1]]*res50$Particles[[1]])), col=c(1,2), lwd=2)

hist(res50$Particles[[26]], freq=FALSE)
abline(v=c(Tstates[[26]], sum(res50$Weights[[26]]*res50$Particles[[26]])), col=c(1,2), lwd=2)

hist(res50$Particles[[50]], freq=FALSE)
abline(v=c(Tstates[[50]], sum(res50$Weights[[50]]*res50$Particles[[50]])), col=c(1,2), lwd=2)


hist(res50$Particles[[100]], freq=FALSE)
abline(v=c(Tstates[[100]], sum(res50$Weights[[100]]*res50$Particles[[100]])), col=c(1,2), lwd=2)
```

